A parallel runtime framework for communication intensive stream applications

### 2013-03-21 paper.tex
kevins last min changes

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Servesh Muralidharan\IEEEauthorrefmark{2},
Kevin Casey\IEEEauthorrefmark{4}
and
David Gregg\IEEEauthorrefmark{2}}
\IEEEauthorblockA{\IEEEauthorrefmark{2}School of Computer Science and Statistics,
Trinity College Dublin, Dublin 2, Ireland\\
Email: muralis@scss.tcd.ie, David.Gregg.cs.tcd.ie }
\IEEEauthorblockA{\IEEEauthorrefmark{4}DCU\\
Email: }
}
%~
%~ \author{\IEEEauthorblockN{Servesh Muralidharan}
%~ \IEEEauthorblockA{School of Computer Science and\\Statistics\\
%~ Trinity College Dublin\\
%~ Dublin 2, Ireland\\
%~ Email: muralis@tcd.ie}
%~ \and
%~ \IEEEauthorblockN{David Gregg}
%~ \IEEEauthorblockA{School of Computer Science and\\Statistics\\
%~ Trinity College Dublin\\
%~ Dublin 2, Ireland\\
%~ Email: David.Gregg@cs.tcd.ie}}

@@

\begin{abstract}
%\boldmath
Stream applications are often limited in their performance by their underlying communication system. In a typical implementation, the underlying communication relies on operating system to handle majority of the network operations. In such cases, the communication stack, which was not designed to handle such tremendous amounts of data, acts as a bottleneck and restricts the performance of the application.

@@

Our runtime system does this by using a series of communication filters that transform the data from the application to packets suitable for transmission. These communication filters are built using system software that allows direct access to network hardware. We also utilize the ability of the modern network controller to classify packets onto multiple hardware queues by integrating it with the communication filter. Using this our runtime framework is able to construct a unified stream graph based on the application graph provided. This when used with a static schedule based on a given architecture, can process the data streams in parallel.

We are able to parallelize the stream application and achieve speedups of more than a factor of eight in all the applications we tested. The results show that our system scales to as many as 24 parallel processes on a multicore computer, and achieves speedups of more than a factor of ten in some cases compared to sequential implementations.

@@

Applications that follow streaming model~\cite{Thies:2002:SLS:647478.727935,Halbwachs91thesynchronous,Stephens95asurvey,muthukrishnan2005data}
are expressed as a network of \textit{filters} connected by communication
\textit{channels}. Tokens of data flow along the communication channels
between filters. Each filter consumes on one or more \textit{streams}
of data from its input channels, and produces streams of data on its
ouput channels. For applications that fit the model well, stream processing
can greatly reduce the working memory required. Since the data flows through
a sequence of filters in a pipelined style, it is often possible to
operate on a relatively small window of the data at any given time,
thereby making it possible to operate on conceptually infinite
data streams.

Traditionally, the stream processing model was used primarily
signal processing type applications, where the application
processes a continuous stream of signal inputs. For example,
many video processing applications are modelled as a set of
image filters operating on a stream of video frame, where the
entire video may be hundreds of gigabytes in size, but the
processing can be done with just a few frames at a time.
However, in more recent years it has been found that the stream
model is highly suitable for many applications that process
very large data sets, in the order of terabytes or even larger.
These include applications such as financial trading
systems~\cite{agarwal2009faster},
database systems~\cite{Golab03datastream},
network data analysis~\cite{Gilbert01quicksand:quick,Sullivan:1998:TSM:1268256.1268258} are few examples that follow the streaming model and operate on very large data sets.

Stream processing is highly suited to multicore and parallel
computing, because the filters can typically execute in parallel.
Smaller stream applications often achieve high levels of parallelism
on multicore machines.  However, as the stream processing model
is used to address larger data processing problems, more computational
resources are needed, and a common solution is to distribute the
filters across a number of machines.
In such cases the applications are typically limited by the
traditional communication
~\cite{Wu:2007:PPB:1323954.1323957,Wu:2007:PAL:1227865.1228021,Dobrescu09routebricks:exploiting,Han:2010:PGS:1851275.1851207,Kohler2000}.
The reason being these systems have limited support for operating on
parallel streams and are burdened by abstraction layers that provide
compatibility for non-streaming applications.

Several high speed networks~\cite{fredj2011survey} exists that could handle such communication requirements but these require expensive hardware and specialized programming skills. Commodity network adapters based on 10-gigabit ethernet (10GbE) are comparatively cheaper alternative and are being widely deployed in clusters and grids in data centers. Previous research\cite{Romanow03anoverview,Feng:2003:OEN:1048935.1050200} have been done on optimizing ethernet for improving communication in such cases. However, they do not use many of the features provided by modern day network adapters such as hardware queues, flow control, etc., Also, they do not focus specifically on streaming applications and hence making it difficult to utilize them properly.

@@

In this paper we propose a parallel processing framework for building stream applications that are bound by very large communication. A simplistic framework where the filters are described as \textit{actors} and the application is represented by connecting such \textit{actors} into a directed acyclic graph (DAG). Then our framework is used to establish, (1)the communication system that allows these \textit{actors} to send and receive data across each other and (2)the computation system that executes these tasks in parallel on a given multicore system. The main contributions can be described as,

\begin{enumerate}
\item Parallel multi point communication layer, which provides the ability for stream actors to communication across the network.
\item Execution system that can run the given data parallel filters on the multicore system
\item Utilizes network hardware features such as queues and flow control to support stream applications

@@

Our main objective is to optimize performance of streaming applications by First, providing a speaclized communication link between the filters and Second, extracting data parallel actors and executing them in parallel. Stream applications usually consists of stateless and stateful filters. Stateful filters are those that are depedendent on the previous execution and hence making it difficult to replicate them. Stateless filters do not pose such limitations and can be freely replicated based on the available data streams. In this paper, we focus on extracting data parallelism based on stateless filters and streams. Example operations of such an application are represented in the stream graph shown in Figure \ref{fig1}. All the filters used in this application are essentially stateless filters. This sample application operates by performing encryption on the data that matches predefined rules like file names, access level of user, etc., otherwise it performs compression. Due to the nature of such stream applications, it is possible to replicate and distribute the filters across different CPUs. We are particularly interested in a scenario where groups of networked servers are used to perform these kind of stream operations by distributing data streams over multiple nodes.

@@

In building a framework for such applications on general purpose computers, we can exploit two hardware features that could allow for dramatically improved performance, namely multicore processors and multi queue network controllers. Using the multiple queues it makes it possible to do much of the communication in parallel and using the multiple cores the filters from the stream application can be executed concurrently.

@@

In our framework, we use a more elegant approach. We define a runtime framework that already consists of the communication filters that are pre-built. So inorder to implement a stream application such as that shown in Figure~\ref{fig1}, all we need to do is define the computation filters and connect them in the framework to the pre-built communication filters. On execution, our runtime framework is able to replicate the stateless filters and the communication filters on to the different multiple cores, thereby executing the operations belonging to different \textit{flows} in parallel.

@@

\comment{Understanding the set of packet processing tasks associated with the application is essential to performing them in parallel.} We use a graph-based representation of filters related to packet processing to determine the dependencies involved for the different stages and to determine the relation with the different tasks of the application. This graph contains the stages through which the data associated with the application is transformed into network packets and vice versa. It shows which of the tasks can be performed in parallel with those of the application, and highlights those that have a data dependency on a specific task in the application. The graph also enables optimizations based on specific characteristics of the application or the architecture used for execution. Figure \ref{fig2} shows a graph representing User Datagram Protocol (UDP) packet generation.

@@

For a communication system to be integrated with an application, it is necessary to have an efficient mechanism for reading and writing packets to and from the NIC without interacting with the operating system. Rapid data access here is crucial for the extraction of data parallelism at later stages. \comment{In using the techniques described below we exploit the modern NIC architecture to its full potential.}

@@

The development of virtualization and improved network flow handling has led to hardware level multi queue support in NICs. During transmission, concurrent writes to different queues are possible, enabling multiple processes to send data simultaneously. During reception, the NIC classifies each packet onto one of the receive queues in a technique known as Receive Side Scaling (RSS)\cite{micro2008} \cite{intel2010}. This technique uses the header information or tuples to classify packets onto different hardware queues. Multi queue support can handle several streams of data in parallel and it is exploited extensively in our framework for this reason. By offloading the classification of packets to the NIC, we remove any overhead associated with packet classification in the application. We can also use these hardware queues as a means of balancing the workload across different threads by controlling the number of queues that each thread handles. \comment{Since different processes can access these transmit and receive queues, concurrent access to data is possible.}

@@

Combining multiple packets that belong to a flow known as \texttt{batching} is a common method used to boost the performance of network applications where several packets are combined and operated upon together, thus reducing the overhead of I/O operations. In our framework, batching is utilized in a form where the number of packets in a batch is varied based on the available space in the hardware queue. This number reflects the amount of data we are able to process before issuing hardware synchronization calls and forms an important means by which we reduce the usage of system calls.

@@

\comment{Computation filters represents the operations associated . The subset of the application that we are interested in parallelizing is the stream processing component.} In stream processing model, the communication flow between tasks is well-defined. \comment{This is usually represented by a synchronous data flow graph (SDF).} The filters associated with the computation can be connected to the communication filters in the same directed acyclic graph (DAG) used to represent the application. \comment{The packet processing tasks operate on packets and the application tasks operate on the data. Since we focus more on parallel packet processing, optimizations in relation to the application graph are not discussed. It is assumed that a static version of the application graph is available and we focus more on executing these tasks in parallel with the packet processing tasks.} The computation filters are designed by the user and are programmed into the runtime framework which already consists of the communication filters.

@@

Two important aspects that the parallel processing engine must address is the scheduling of filters and balancing the load across the available resources. The parallel processing engine consists of collaborating processes which are bound to specific CPUs and act as engines to execute the workload. Based on their requirements and the available resources, filters are scheduled across these execution engines. We assume a static schedule of the filters on to the given multiple cores is provided for execution. \comment{In this manner, the workload can be distributed flexibly.}

@@

\comment{Most extant stream applications depend on the OS for their communication tasks.}In commodity systems, the interaction with the OS for network operations is a necessary abstraction to handle multiple applications. However, in the case of specialized servers which cater to specific applications, it would be beneficial to interact directly with the network device. Moreover, in the case of intensive network traffic, the OS could act as a bottleneck due to the overhead associated with data flowing through the kernel before being transmitted through the network device\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}. Eliminating this would be possible, either by executing the application in the kernel space (which could compromise stability of the system), or by using an interface for accessing the NIC buffers from user space. The latter approach is supported by several interfaces such as \texttt{netmap} \cite{Rizzo:2012:RNI:2090147.2103536}, PF\_RING DNA \cite{1564468}, UIO-IXGBE \cite{Krasnyansky}, etc., Even though these interfaces support user space packet access, they lack the ability to be invoked in a efficient manner by the application. More specifically, they do not have any way of optimizing the rate of processing, lacking, for example, the concept of flows for application tasks. The application tasks are essentially the computation filters defined in our design principles. The lack of support for interacting with these interfaces in order to read and write data in parallel from concurrent processes or threads adds to the problem. Also, since the application is now responsible for handling network packets, it has to perform the additional operations that would otherwise be carried out by the OS.

@@

\subsubsection{Userspace Packet Access}
Applications executed in kernel space for accessing hardware devices directly would lead to system instability. Instead, previous researchers \cite{Rizzo:2012:RNI:2090147.2103536}, \cite{1564468}, \cite{Krasnyansky}, etc., propose access to network buffers from user space. We use \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} to access data from the network buffers. \texttt{Netmap} uses a set of ring buffers in kernel space and user space to export the network hardware buffer's memory region. This enables the applications in user space to directly access the network hardware. It also prevents the application from accessing memory outside valid regions by checking the limits of the ring buffer. Packets can be written to the buffers when space is available and an \textit{ioctl} call to the driver enables the network controller to send the data. A \textit{poll} interface is implemented on the memory region mapped in user space to check when new packets are available. On reading or writing packets, the available space is updated.

@@

\item Provide batching of packets based on available space on hardware rings
\item Reduce the usage of system calls such as \textit{ioctl} and \textit {poll}
\item Provide mechanisms for handling multiple queues through the application

@@

Algorithms \ref{s_pack} and \ref{r_pack} use a batch update for sending and receiving packets based on the rate at which data flows through the interface. This is done by maintaining a data structure that has information about the hardware rings, updating it on a periodic basis. Once a fixed amount of space is determined, we update this on our local structure. We then continue to read or write data from the hardware ring, but update the actual limits and position on our local structure. When we have used up the space that was allocated, we update the hardware ring with the changes. In cases where the application is generating sparse traffic, it is possible to frequently update the ring in order to maintain a consistent latency in transmission. Multi queue support is provided by managing the process of selecting the ring to send or receive based on an application determined function. The $f(n)$ present in Algorithm \ref{s_pack} line \ref{s_pack_param} and Algorithm \ref{r_pack} line \ref{r_pack_param} represents this. Choosing a random queue ($f(n)=X \sim U([0,n])$ n=no of queues)has proven to be effective in our experiments, but if an application requires a specific flow to be maintained, a different function here could provide that support. This would be helpful in cases where the data operations are more predictable, or if the transactions seem to follow a specific pattern. The $threshold$ specified in Algorithm \ref{r_pack} line \ref{r_pack_thres} is determined for each application based on the rate at which it is able to process data. By issuing the system calls such as \textit{ioctl} and \textit{poll} only when required, and effectively handling the multiple queues, we are able to achieve improved performance.

@@

In order to evaluate our framework, we test the scalability and the performance speedup on a multicore system for an application that is constructed using the framework. Using two generic operations: compression and encryption, we build simple stream applications using our framework. We then measure the performance improvement as we increase the number of execution engines dedicated to the application.

@@

Each application is constructed as two parts, one a source of data and the other a sink. We measure the overall time it takes for application task operations, communication operations, and the time to send from the source and receive at the sink. This is done for different packet sizes and applications. Then we plot the speedup achieved in each of the cases over the serial version where we have a single sender and receiver operating on the multiple queues. This is the baseline in our experiments and is similar to how the application would behave if it were implemented using just direct access to the network hardware.

@@

Both applications are constructed similarly. Data sizes of 1.28GB and 2.56GB are selected as the workload for the compression and encryption applications respectively. The OpenSSL library is used for encryption/decryption and zlib is used for compression/decompression operations. Each application consists of a data generation phase, then data is compressed or encrypted depending on the application. Packet operations are performed after the data is generated. The stream consisting of packets is then sent over the network using multiple queues. The sink side performs the reverse of these operations. Once data is received, packet operations are performed to extract the payload, followed by decompression or decryption to retrieve the original data. These are representative of some of the basic operations that would be involved in a much more complex application. \comment{Choosing to compare these simple operations avoids any extraneous factors that may be present in a more complex application.}Use of these simpler stream applications enables us to evaluate the performance of our framework and not that of any optimizations specific to the application.

@@

One is that, from fig\ref{fig-res}.(d,e,f), we note that compression requires more work than decompression, whereas from fig\ref{fig-res}.(a,b,c) encryption and decryption are more balanced. This is evident from the continued speedup we get up to 4 senders in the case of compression/decompression. Only a slight asymmetry can be seen in the encryption/decryption case, where adding more senders than receivers only yields a slight increase in performance. In this case, the likely explanation is support for AES instructions in the architecture, making decryption a marginally easier task. In general though, the performance here only improves by increasing both the number of senders and receivers simultaneously. Even from these preliminary results, it is clear that the optimal sender/receiver balance varies from application to application and highlights the need for effective load-balancing.

@@

Stream programming languages such as StreamIt~\cite{Thies:2002:SLS:647478.727935}, LUSTRE~\cite{Halbwachs91thesynchronous} and extensions such as Carpenter et al.~\cite{Carpenter:2007:SMD:1776200.1776217}, Buck et al.~\cite{Buck:2004:BGS:1015706.1015800} target stream applications but the focus is more on parallelizing the computations related to the application alone on the multicore system unlike our runtime framework which parallelizes communication operations also.

MPICH-G2~\cite{Karonis02mpich-g2:a} targets large grid networks and for MPI programming model unlike us which is focused more on streaming model. Moreover, we do not employ any of the MPI communication primitives provided and use our own runtime system to communicate data between the filters in the stream graph across systems.

Wagner et al.~\cite{5160944} propose a stream processing framework that completely utilizes the MPI library. They make use of MPI groups and communicators to improve the flexibility of MPI library to support stream-processing. Our runtime framework is designed to provide a low overhead and parallel communication framework by directly utilizing the network link. Whereas they employ a library using MPI primitives to construct their stream-processing system and rely on the underlying abstraction layer to communicate over any network the MPI library supports.

@@

While we use MPI to execute the computation filters, our runtime framework employs a direct and parallel communication interface for the actual transmission of reception of data between filters. This is the key feature that enables us to parallelize the communication along with the computation for a given streaming application, which is essential in improving its performance.

@@

\section{Conclusions}
In this paper we have proposed a parallel runtime framework that can integrate communication operations with the stream applications and perform both types of operation in parallel.

Our algorithms for sending and receiving data, which implicitly perform batching and reduce the usage of system calls, and by combining this with \texttt{netmap} APIs which minimizes the interaction with the OS, communication performance in our framework is able to scale with available processors. Instead of performing the communication operations separately in the OS, by integrating them with the stream application \comment{using Click's elements within our infrastructure,} we are able to construct a unified stream graph that represents the computation and communication filters. Using this, we are able to parallelize the stream application and achieve speedups of more than a factor of eight in all the applications we tested.

In integrating these features into a runtime framework, the system of multiple queues, parallel communication and computation are entirely hidden from the programmer, who merely specifies a standard stream graph for computation alone. \comment{This frees up the programmer from the onerous task of having to decide how their application should be distributed, leaving it instead to the execution engine of our framework, which also looks after the ensuing communication requirements after application distribution.}

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Servesh Muralidharan\IEEEauthorrefmark{2},
Kevin Casey\IEEEauthorrefmark{4}
and
David Gregg\IEEEauthorrefmark{2}}
\IEEEauthorblockA{\IEEEauthorrefmark{2}School of Computer Science and Statistics,
Trinity College Dublin, Dublin 2, Ireland\\
Email: muralis@scss.tcd.ie, David.Gregg@cs.tcd.ie }
\IEEEauthorblockA{\IEEEauthorrefmark{4}School of Computing, Dublin City University, Glasnevin, Dublin 9, Ireland\\
Email: Kevin.Casey@computing.dcu.ie}
}
%~
%~ \author{\IEEEauthorblockN{Servesh Muralidharan}
%~ \IEEEauthorblockA{School of Computer Science and\\Statistics\\
%~ Trinity College Dublin\\
%~ Dublin 2, Ireland\\
%~ Email: muralis@tcd.ie}
%~ \and
%~ \IEEEauthorblockN{David Gregg}
%~ \IEEEauthorblockA{School of Computer Science and\\Statistics\\
%~ Trinity College Dublin\\
%~ Dublin 2, Ireland\\
%~ Email: David.Gregg@cs.tcd.ie}}

@@

\begin{abstract}
%\boldmath
Stream applications are often limited in their performance by their underlying communication system. A typical implementation relies on the operating system to handle the majority of network operations. In such cases, the communication stack, which was not designed to handle such tremendous amounts of data, acts as a bottleneck and restricts the performance of the application.

@@

Our runtime system does this by using a series of communication filters that transform the data from the application to packets suitable for transmission. These communication filters are built using system software that allows direct access to network hardware. We also utilize the ability of the modern network controller to classify packets onto multiple hardware queues by integrating it with the communication filter. Using this, our runtime framework is able to construct a unified stream graph based on the application graph provided. This, when used with a static schedule based on a given architecture, can process the data streams in parallel.

We are able to parallelize stream applications and achieve speedups of more than a factor of eight in all the applications we tested. The results show that our system scales to as many as 24 parallel processes on a multicore computer, and achieves speedups of more than a factor of ten in some cases compared to sequential implementations.

@@

Applications that follow a streaming model~\cite{Thies:2002:SLS:647478.727935,Halbwachs91thesynchronous,Stephens95asurvey,muthukrishnan2005data}
are expressed as a network of \textit{filters} connected by communication
\textit{channels}. Tokens of data flow along the communication channels
between filters. Each filter consumes on one or more \textit{streams}
of data from its input channels, and produces streams of data on its
ouput channels. For applications that fit the model well, stream processing
can greatly reduce the working memory required. Since the data flows through
a sequence of filters in a pipelined style, it is often possible to
operate on a relatively small window of the data at any given time,
thereby making it possible to operate on conceptually infinite
data streams.

Traditionally, the stream processing model was used primarily
for signal processing type applications, where the application
processes a continuous stream of signal inputs. For example,
many video processing applications are modelled as a set of
image filters operating on a stream of video frame, where the
entire video may be hundreds of gigabytes in size, but the
processing can be done with just a few frames at a time.
However, in more recent years it has been found that the stream
model is highly suitable for many applications that process
very large data sets, in the order of terabytes or even larger.
Applications such as financial trading
systems~\cite{agarwal2009faster},
database systems~\cite{Golab03datastream},
network data analysis~\cite{Gilbert01quicksand:quick,Sullivan:1998:TSM:1268256.1268258} are a few examples that follow the streaming model and also operate on very large data sets.

Stream processing is highly suited to multicore and parallel
computing, because the filters can typically execute in parallel.
Smaller stream applications often achieve high levels of parallelism
on multicore machines.  However, as the stream processing model
is used to address larger data processing problems, more computational
resources are needed, and a common solution is to distribute the
filters across a number of machines.
In such cases, the applications are typically limited by the
traditional communication
~\cite{Wu:2007:PPB:1323954.1323957,Wu:2007:PAL:1227865.1228021,Dobrescu09routebricks:exploiting,Han:2010:PGS:1851275.1851207,Kohler2000}.
The main reason for this is that these systems have limited support for operating on
parallel streams and are burdened by abstraction layers that provide
compatibility for non-streaming applications.

Several high speed networks~\cite{fredj2011survey} exist that could handle such communication requirements but these require expensive hardware and specialized programming skills. Commodity network adapters based on 10-gigabit ethernet (10GbE) are a comparatively cheaper alternative and are being widely deployed in clusters and grids in data centers. Previous research\cite{Romanow03anoverview,Feng:2003:OEN:1048935.1050200} has been done on optimizing ethernet for improving communication in such cases. However, they do not use many of the features provided by modern day network adapters such as hardware queues, flow control, etc., Also, they do not focus specifically on streaming applications and hence making it difficult to utilize them properly.

@@

In this paper we propose a parallel processing framework for building stream applications that are bound by very large communication. A simple framework is used, where the filters are described as \textit{actors} and the application is represented by connecting such \textit{actors} into a directed acyclic graph (DAG). Then our framework is used to establish, (1)the communication system that allows these \textit{actors} to send and receive data across each other and (2)the computation system that executes these tasks in parallel on a given multicore system. The main contributions can be described as,

\begin{enumerate}
\item A parallel multi point communication layer, which provides the ability for stream actors to communicate across the network
\item An execution system that can run the given data-parallel filters on the multicore system
\item The utilization of network hardware features such as queues and flow control to support stream applications.

@@

Our main objective is to optimize performance of streaming applications by first, providing a specialized communication link between the filters and second, extracting data-parallel actors and executing them in parallel. Stream applications usually consists of stateless and stateful filters. Stateful filters are those that are depedendent on the previous execution which makes it difficult to replicate them. Stateless filters do not pose such limitations and can be freely replicated based on the available data streams. In this paper, we focus on extracting data parallelism based on stateless filters and streams. Example operations of such an application are represented in the stream graph shown in Figure \ref{fig1}. All the filters used in this application are essentially stateless filters. This application operates by performing encryption on the data that matches predefined rules like file names, access level of user, etc., otherwise it performs compression. Due to the nature of such stream applications, it is possible to replicate and distribute the filters across different CPUs. We are particularly interested in a scenario where groups of networked servers are used to perform these kind of stream operations by distributing data streams over multiple nodes.

@@

In building a framework for such applications on general purpose computers, we can exploit two hardware features that could allow for dramatically improved performance, namely multicore processors and multi queue network controllers. Using the multiple queues makes it possible to do much of the communication in parallel and by using the multiple cores, the filters from the stream application can be executed concurrently.

@@

In our framework, we use a more elegant approach. We define a runtime framework that provides pre-built communication filters. In order to implement a stream application such as that shown in Figure~\ref{fig1}, all we need to do is define the computation filters and connect them in the framework to the pre-built communication filters. On execution, our runtime framework is able to replicate the stateless filters and the communication filters onto the different multiple cores, thereby executing the operations belonging to different \textit{flows} in parallel.

@@

\comment{Understanding the set of packet processing tasks associated with the application is essential to performing them in parallel.} We use a graph-based representation of filters related to packet processing to determine the dependencies involved for the different stages and to determine the relation with the different tasks of the application. This graph contains the stages through which the data associated with the application is transformed into network packets and vice versa. It shows which of the tasks can be performed in parallel with those of the application, and highlights those that have a data dependency on a specific task in the application. The graph also enables optimizations based on specific characteristics of the application or the architecture used for execution. Figure \ref{fig2} shows a simple graph representing User Datagram Protocol (UDP) packet generation.

@@

For a communication system to be integrated with an application, it is necessary to have an efficient mechanism for reading and writing packets to and from the NIC without interacting with the operating system. Rapid data access here is crucial for the extraction of data parallelism at later stages. We use a number of techniques to support this rapid data access.\comment{In using the techniques described below we exploit the modern NIC architecture to its full potential.}

@@

The development of virtualization and improved network flow handling has led to hardware-level multi queue support in NICs. During transmission, concurrent writes to different queues are possible, enabling multiple processes to send data simultaneously. During reception, the NIC classifies each packet onto one of the receive queues in a technique known as Receive Side Scaling (RSS)\cite{micro2008} \cite{intel2010}. This technique uses the header information or tuples to classify packets onto different hardware queues. Multi queue support can handle several streams of data in parallel and it is exploited extensively in our framework for this reason. By offloading the classification of packets to the NIC, we remove any overhead associated with packet classification in the application. We can also use these hardware queues as a means of balancing the workload across different threads by controlling the number of queues that each thread handles. \comment{Since different processes can access these transmit and receive queues, concurrent access to data is possible.}

@@

A common technique known as \texttt{batching} is used to boost the performance of network applications, where several packets are combined and operated upon together, thus reducing the overhead of I/O operations. In our framework, batching is utilized in a form where the number of packets in a batch is varied based on the available space in the hardware queue. This number reflects the amount of data we are able to process before issuing hardware synchronization calls and forms an important means by which we reduce the usage of system calls.

@@

\comment{Computation filters represents the operations associated . The subset of the application that we are interested in parallelizing is the stream processing component.} In stream processing model, the communication flow between tasks is well-defined. \comment{This is usually represented by a synchronous data flow graph (SDF).} The filters associated with the computation can be connected to the communication filters in the same directed acyclic graph (DAG) used to represent the application. \comment{The packet processing tasks operate on packets and the application tasks operate on the data. Since we focus more on parallel packet processing, optimizations in relation to the application graph are not discussed. It is assumed that a static version of the application graph is available and we focus more on executing these tasks in parallel with the packet processing tasks.} The computation filters are designed by the user and are programmed into the runtime framework which already provides the communication filters.

@@

Two important aspects that the parallel processing engine must address is the scheduling of filters and the balancing of the load across the available resources. The parallel processing engine consists of collaborating processes which are bound to specific CPUs and act as engines to execute the workload. Based on their requirements and the available resources, filters are scheduled across these execution engines. We assume a static schedule of the filters onto the given multiple cores is provided for execution. \comment{In this manner, the workload can be distributed flexibly.}

@@

\comment{Most extant stream applications depend on the OS for their communication tasks.}In commodity systems, the interaction with the OS for network operations is a necessary abstraction to handle multiple applications. However, in the case of specialized servers which cater to specific applications, it would be beneficial to interact directly with the network device. Moreover, in the case of intensive network traffic, the OS could act as a bottleneck due to the overhead associated with data flowing through the kernel before being transmitted through the network device\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}. Eliminating this would be possible, either by executing the application in the kernel space (which could compromise stability of the system), or by using an interface for accessing the NIC buffers from user space. The latter approach is supported by several interfaces such as \texttt{netmap} \cite{Rizzo:2012:RNI:2090147.2103536}, PF\_RING DNA \cite{1564468}, UIO-IXGBE \cite{Krasnyansky}, etc., Even though these interfaces support user space packet access, they lack the ability to be invoked in a efficient manner by the application. More specifically, they do not have any way of optimizing the rate of processing, lacking, for example, the concept of flows for application tasks. These application tasks are represented as computation filters similar to those described in figure~\ref{fig1}. The lack of support for interacting with these interfaces in order to read and write data in parallel from concurrent processes or threads adds to the problem. Also, since the application is now responsible for handling network packets, it has to perform the additional operations that would otherwise be carried out by the OS.

@@

\subsubsection{Userspace Packet Access}
Applications executed in kernel space for accessing hardware devices directly would lead to system instability. Instead, previous researchers \cite{Rizzo:2012:RNI:2090147.2103536}, \cite{1564468}, \cite{Krasnyansky}, etc., propose access to network buffers from user space. We use \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} to access data from the network buffers. \texttt{Netmap} uses a set of ring buffers in kernel space and user space to export the network hardware buffer's memory region. This enables applications in user space to directly access the network hardware. It also prevents applications from accessing memory outside valid regions by checking the limits of the ring buffer. Packets can be written to the buffers when space is available, and an \textit{ioctl} call to the driver enables the network controller to send the data. A \textit{poll} interface is implemented on the memory region mapped in user space to check when new packets are available. On reading or writing packets, the available space is updated.

@@

\item Provide batching of packets based on available space on hardware rings
\item Reduce the usage of system calls such as \textit{ioctl} and \textit {poll}
\item Provide mechanisms for handling multiple queues through the application.

@@

Algorithms \ref{s_pack} and \ref{r_pack} use a batch update for sending and receiving packets based on the rate at which data flows through the interface. This is done by maintaining a data structure that has information about the hardware rings, updating it on a periodic basis. Once a fixed amount of space is determined, we update this on our local structure. We then continue to read or write data from the hardware ring, but update the actual limits and position on our local structure. When we have used up the space that was allocated, we update the hardware ring with the changes. In cases where the application is generating sparse traffic, it is possible to frequently update the ring in order to maintain a consistent latency in transmission. Multi queue support is provided by managing the process of selecting the ring to send or receive based on an application determined function. The $f(n)$ present in Algorithm \ref{s_pack} line \ref{s_pack_param} and Algorithm \ref{r_pack} line \ref{r_pack_param} represents this. Choosing a random queue ($f(n)=X \sim U([0,n])$ n=no of queues) has proven to be effective in our experiments, but if an application requires a specific flow to be maintained, a different function here could provide that support. This would be helpful in cases where the data operations are more predictable, or if the transactions seem to follow a specific pattern. The $threshold$ specified in Algorithm \ref{r_pack} line \ref{r_pack_thres} is determined for each application based on the rate at which it is able to process data. By issuing the system calls such as \textit{ioctl} and \textit{poll} only when required, and effectively handling the multiple queues, we are able to achieve improved performance.

@@

In order to evaluate our framework, we test the scalability and the performance speedup on a multicore system for an application that is constructed with the framework. Using two generic operations: compression and encryption, we build simple stream applications. We then measure the performance improvement as we increase the number of execution engines dedicated to the application.

@@

Each application is constructed in two parts, one a source of data and the other a sink. We measure the overall time it takes for application task operations, communication operations, and the time to send from the source and receive at the sink. This is done for different packet sizes and applications. Then we plot the speedup achieved in each of the cases over the serial version where we have a single sender and receiver operating on the multiple queues. This is the baseline in our experiments and is similar to how the application would behave if it were implemented using just direct access to the network hardware.

@@

Both applications are constructed similarly. Data sizes of 1.28GB and 2.56GB are selected as the workload for the compression and encryption applications respectively. The OpenSSL library is used for encryption/decryption and zlib is used for compression/decompression operations. Each application consists of a data generation phase, then data is compressed or encrypted depending on the application. Packet operations are performed after the data is generated. The stream consisting of packets is then sent over the network using multiple queues. The sink side performs the reverse of these operations. Once data is received, packet operations are performed to extract the payload, followed by decompression or decryption to retrieve the original data. These are representative of some of the typical operations that would be involved in a much more complex application. \comment{Choosing to compare these simple operations avoids any extraneous factors that may be present in a more complex application.}Use of these simpler stream applications enables us to evaluate the performance of our framework and not that of any optimizations specific to the application.

@@

One is that, from fig\ref{fig-res}.(d,e,f), we note that compression requires more work than decompression, whereas from fig\ref{fig-res}.(a,b,c) encryption and decryption are more balanced. This is evident from the continued speedup obtained from up to 4 senders in the case of compression/decompression. Only a slight asymmetry can be seen in the encryption/decryption case, where adding more senders than receivers only yields a slight increase in performance. In this case, the likely explanation is support for AES instructions in the architecture, making decryption a marginally easier task. In general though, the performance here only improves by increasing both the number of senders and receivers simultaneously. Even from these preliminary results, it is clear that the optimal sender/receiver balance varies from application to application and highlights the need for effective load-balancing.

@@

Stream programming languages such as StreamIt~\cite{Thies:2002:SLS:647478.727935}, LUSTRE~\cite{Halbwachs91thesynchronous} and extensions such as Carpenter et al.~\cite{Carpenter:2007:SMD:1776200.1776217}, Buck et al.~\cite{Buck:2004:BGS:1015706.1015800} target stream applications, but the focus is more on parallelizing the computations related to the application alone on the multicore system unlike our runtime framework which also parallelizes communication operations.

MPICH-G2~\cite{Karonis02mpich-g2:a} targets large grid networks and applications written using the MPI programming model, unlike our focus on the streaming model. Moreover, we do not employ any of the MPI communication primitives provided to handle streaming data. Instead we use our own fast runtime system to communicate data between the filters in the stream graph across systems.

Wagner et al.~\cite{5160944} propose a stream processing framework that completely utilizes the MPI library. They make use of MPI groups and communicators to improve the flexibility of the MPI library to support stream-processing. Their approach employs a library using MPI primitives to construct their stream-processing system and relies on the underlying abstraction layer to communicate over any network the MPI library supports. Our runtime framework is designed to provide a low overhead, parallel communication framework by directly utilizing the network link.

@@

While we use MPI to execute the computation filters, our runtime framework employs a direct and parallel communication interface for the actual transmission and reception of data between filters. This is the key feature that enables us to parallelize the communication along with the computation for a given streaming application, which is essential in improving its performance.

@@

\section{Conclusions}
In this paper we have proposed a parallel runtime framework that can integrate communication and computational operations in stream applications and perform both types of operation in parallel.

Our algorithms for sending and receiving data implicitly perform batching and reduce the usage of system calls, and we combine this with \texttt{netmap} APIs to minimize the interaction with the OS. This allows communication performance in our framework to better scale with the number of available processors. Instead of performing the communication operations separately in the OS, by integrating them within the stream application\comment{using Click's elements within our infrastructure,}, we are able to construct a unified stream graph that represents the entire application, including computation and communication filters. Using this, we are able to parallelize these stream applications and achieve speedups of more than a factor of eight in all the applications we tested.

By integrating these features into a runtime framework, the system of multiple queues, parallel communication and computation are entirely hidden from the programmer, who merely specifies a standard stream graph for computation alone. This frees up the programmer from the onerous task of dealing with the substantial communication requirements that result from stream application distribution.
